Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-03-01 draft.tex
.

\begin{abstract}
Building upon on a self-organizing spiking neural network (LIF-SORN), we attempted to replace the intrinsic homeostatic control system used in earlier versions by a mechanism based on the diffusion of a neurotransmitter across the nervous tissue. The model of diffusive homeostasis was adopted from a paper by Sweeney et al. and models the tissue as a surface of square shape and a set of points on this surface, representing the positions of the neurons within the network. Excitatory neurons acted as a source of nitric oxide (NO), as well as as a sensor for the NO concentration at each individual position. Production and sensing of NO formed the basis of a feedback loop: Individual NO readouts caused an appropriate change within the internal excitability of the neurons. The control system was closed by linking the rate of NO production to the neurons' firing rates.

The main goal of this modification was to allow for a broad and heavy tailed distribution of firing rates among the excitatory neural population. This feature of cortical activity has been extensively observed and studied experimentally, but could not be achieved by the formerly used single-cell homeostatic mechanism which bound firing rates of all neurons to a fixed target value. Indeed, our statistical analysis of spiking activity showed that diffusive homeostasis could significantly broaden the distribution of firing rates.

@@

\section{Introduction}
\subsection{Homeostasis in the Nervous System}
Among other functional roles of the mammalian central nervous system it is involved in maintaining crucial properties of the body in a functional regime. Some examples regarding the organism as a whole are e.g. body temperature and blood pressure, both being controlled by hypothalamus \cite[p. 484]{Bear_Exploring_the_Brain}. The term \textit{homeostasis} entailing these regulatory mechanisms was first introduced by Walter C. Cannon in 1932 \cite{Wisdom_of_the_Body_Cannon, Homeostasis_Queenan_2012}. Experimental studies have found evidence for processes within the brain that do not function as a regulatory control for certain aspects of the body, but as a homeostatic control of neural activity itself \cite{Turrigiano_1994,Burrone_2003}in more recent years. Many processes underlying this self-regulatory behavior have been found up till now and many different biophysical interdependencies have shown to play a role. This diversity makes comparability of these mechanisms a rather difficult task. On a relatively general level though experiments have revealed two main types of neuronal regulation: synaptic homeostatic plasticity and intrinsic homeostasis \cite{Desai_2003}.

The notion of synaptic homeostatic plasticity subsumes the observation that neurons - being more or less active than a certain set point of mean activity - are regulated by means of their incoming synaptic weights. Theoretical and experimental studies have suggested that this type of regulation is taking place in a multiplicative manner, thereby retaining proportions between individual synaptic strengths \cite{Syn_Plast_Abbott}. One can regard this form of plasticity as an antagonist to other aspects of synaptic plasticity that affect synaptic strengths in an activity dependent, hebbian fashion. Forms of regulation belonging to intrinsic homeostasis alter the neurons' capability of spiking upon external input by changing intrinsic properties of the nerve cell. More precisely, a neurons' ability to form an action potential can change by modifying the conductance of ion channels in its cell membrane \cite[p. 156]{Theor_Neur_Dayan}.

Theoretical models of homeostasis are generally based on a form of feedback control.This is also known as a \textit{closed loop control} in control theory. All these systems include a number of necessary functional compartments, which will be briefly introduced in the following. Concepts and terms were taken from \cite[p. 59-64]{Cybernetik_Systems_Cruse_2006}.
\newpage
\subsubsection{Theory of Control Systems} \label{Control_Sys_Theo_Section}
A closed loop control generally must contain some \emph{process} that is to be regulated. Despite its potential internal complexity, its functional role in the context of a feedback system is to simply generate a time-dependent output based on the history of a time-dependent input. This output is what one wishes to control and regulate by means of the feedback loop and thus represents the signal that is further processed. It might not be possible however to directly quantify this desired output in terms of some internal state of the process. The firing rate of a neuron for example is an abstract quantity, which can not be \emph{directly} related to some neuronal state variable. Consequently, it might be necessary to transform the actual output of the process into a different signal by a \textit{feedback transducer}.The (potentially transformed) output is fed into a \textit{comparator} in the next step, which - unsurprisingly - compares the given output of the process to some given target or \textit{reference}, typically by subtraction. Note that this reference can but does note have to be a time-dependent signal. The term \textit{homeostasis} however generally refers to a fixed reference, while implementations with a time dependent reference can be regarded as a form of \textit{servomechanism}. The difference between reference and the feedback transducer's signal is known as the \textit{error} signal. This error signal is given to a \textit{controller}, which transforms this error signal into an appropriate input signal for the initially introduced process. This is also the point in the control loop where \textit{external disturbances} come into play by being added to the input generated by the controller. We illustrated the described flow of feedback control in Figure \ref{Closed_Loop_Control_Illustration}.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/illustrations/closed_loop_control.eps}
\end{center}
\caption{Schematic illustration of a closed loop control as described in Section \ref{Control_Sys_Theo_Section}}
\label{Closed_Loop_Control_Illustration}
\end{figure}
Based on the way the controller generates input for the dynamic process based on the error signal one can classify control loops by three types:
\newpage
\begin{itemize}
\item A \textit{proportional controller} generates a signal that is proportional to the error signal it receives. The most important property of a proportional controller is that the stationary state of the control loop generally does not imply an error signal equal to zero.
\item An \textit{integrator controller} uses an integration over the past of the error signal to obtain its output. The stationary solution necessarily implies that the error signal is zero.
\item A \textit{differential controller} generates its output by means of the time derivative of the error signal. This type of controller can quickly react to momentary changes but does not compensate for sustained deviations of the error signal from zero.
\end{itemize}
It should be pointed out that these types of control mechanisms can be simultaneously used in a control system, e.g. through a linear combination. Furthermore, the presented terms refer to concepts of linear control theory. In the study of nonlinear feedback systems they thus only apply in a limited regime where a first order approximation of the system is sufficiently exact.

\subsubsection{Error Signals and Controllers of Neuronal Homeostasis}
A key step in characterizing a control loop in the context of biological systems should be to identify the previously described compartments. While the functional separation illustrated in Figure \ref{Closed_Loop_Control_Illustration} might seem intuitive and ``straight forward", it is not obvious which physiological element should be attributed to a certain abstract element.
As a start, we state that the process and its output controlled by homeostasis is the generation of action potentials and the resulting firing rate. Firing rate is - as stated earlier - an abstract quantity, requiring a feedback transducer to translate the neuron's spiking activity into a ``sensible" quantity. In a nerve cell this could be represented by e.g. the relation between the firing of action potentials and the temporary rise in $\mathrm{Ca^{2+}}$ concentration. An example for a homeostatic model making use of intracellular calcium concentration as a measure of spiking activity is the LeMasson-Abbott model \cite{LeMasson_1993}. More complex chains of activity encoding are of course possible and we will introduce a recent model making additional use of the concentration of nitric oxide in Section \ref{DIff_Hom_Intro_Section} and \ref{NO_Experiments_Section}. Given a representation of activity, its deviation from a reference value will cause other physiological compartments to affect the neuron's ability to generate action potentials. Referring to the concepts introduced in the previous section one could then ask what type of control best resembles the actual neuronal control. The fact that neurons, whose firing rate was artificially increased or suppressed over a long time, can reliably tune their activity back to the initial state (see e.g. \cite{Burrone_2002}) rules out a purely differential type of control. The return to baseline under different disturbances seems to point towards an integral-like control. It is clear however that a the ability of a biological system to encode the accumulation of an error signal is limited by the finite amount of available resources. The conductance of an ion channel for example has natural upper and lower bounds that it can not surpass. We will pick up this issue in Section \ref{Possible_Modifications_Section}.

\subsubsection{The Role of Firing Rate Heterogeneity -  a Dichotomy between Heterogeneity and Control?}
Modeling homeostasis necessarily includes a form of error estimation, which requires setting some predefined target that the feedback control will attempt to achieve. Of course this raises the question of how to choose these homeostatic targets. Experimental studies have found that neuronal populations exhibit a strong heterogeneity within their individual activity. Many statistical analyses found particular evidence for a log-normal like distribution of firing rates \cite{Buzsaki_Fir_Rates_2014,Wohrer_Fir_Rates_2012}. Following these results, one approach of modeling this feature would be to randomly distribute individual targets according to the aforementioned statistics. A problem of this procedure is the absence of any particular deeper physiological motivation or causation. In the light of this conceptual problem, one could, due to a lack of better knowledge, simply define a single global target of activity.

However, numerous experimental and theoretical studies have suggested that the presence of both slow- and fast-firing cells is not to be regarded as an ignorable, mere side effect of brain dynamics. It rather seems to be that skewed and heavy-tailed firing rate statistics are necessary with respect to the preservation of stable synaptic structures, being related to long-term memory, while still allowing synaptic rewiring in order to adapt to changes in external stimuli. For example, experiments by Dragoi et al. have found that during remapping of hippocampal place cells, neurons with low firing rates are subject to stronger changes within their field of spatial representation than those with high firing rates \cite{Dragoi_2003}. In this context Buzsáki and Mizuseki argue that a small subset of highly active neurons may serve to form relatively strongly connected and stable sub-networks that can quickly react to momentary changes in the environment, while fine tuned adaptation involves the slower firing majority of cells \cite{Buzsaki_Fir_Rates_2014}. A realistic network model of neural activity and plasticity should therefore allow for a broad distribution of firing rates to cope with these functional aspects.

@@

Another problem that one might consider is the fact that temporal fluctuations of activity are believed to be a major part of information transmission and processing within the brain \cite[p. 123-150]{Theor_Neur_Dayan}. Homeostasis as described above could potentially override information by eliminating these fluctuations of activity. However, it is argued that homeostatic mechanisms act on time scales of hours to days and thus do not interfere with changes in activity taking place on the order of seconds or below \cite{Turrigiano_2011}. Returning to the previous question, though the slowness of homeostatic adaptation might allow for temporal fluctuations of activity, it does not overcome the conceptual dichotomy between single cell homeostasis and persistent variability in spontaneous activity. Marder and Goaillard speculated that instead of a mechanism controlling activity on a single-cell level, one might consider the possibility of a ``global sensor of network performance to provide error signals for tuning of the network" \cite{Marder_2006}. This concept was used by Sweeney et al. as the basic idea for the introduction of a diffusive, intercellular homeostatic mechanism, which will be introduced in the next section.

\subsection{Diffusive Homeostasis}\label{DIff_Hom_Intro_Section}
In this thesis we examined \textit{diffusive homeostasis} as a possible candidate to replace a previously used model of single-cell intrinsic plasticity. The idea and modeling of diffusive homeostasis was adopted from a paper by Sweeney et al. \cite{Sweeney_Paper}, which models neural tissue as a two-dimensional surface and a set of points representing the neurons' positions. The group of excitatory neurons acted as a point-source of nitric oxide (NO) as well as as a sensor for the NO-concentration at each individual position. The production and sensing of NO forms the basis of a feedback loop: The individual NO-readout is fed into a comparator which causes an appropriate change within the internal firing threshold of the neuron, in turn altering the neuron's firing rate. The control system is then closed by linking the rate of NO-production to the neuron's firing rate. The results by Sweeney et al. suggest that diffusive signaling could resolve the dichotomy between overall stability of firing activity and the need to allow flexibility among individual neurons' activities. Through the diffusive signal, each neurons receives - intuitively speaking - a mixture of its own activity and its neighboring neurons. Individual tuning of firing rates is thereby suppressed while the overall population activity is kept at a constant level.

\subsubsection{Nitric Oxide in the Nervous System}\label{NO_Experiments_Section}
The model presented in \cite{Sweeney_Paper} includes spatial interaction across excitatory neurons through the diffusion of nitric oxide (NO). Enzymes that are responsible for NO synthesis (NOS) are present in different areas of the body, namely being involved in the dilation of blood vessels (endothelial NOS, eNOS), the immune system (inducible NOS, iNOS) and the CNS (neuronal NOS, nNOS) \cite{NOS_Mammals}. These Enzymes differ in their functionality and dependence on the presence of other molecules. In particular, nNOS is sensitive to the concentration of $Ca^{2+}$ ionic Calcium \cite{Knowles_Ca_nNOS,Steinert_NO}. Nitric oxide that was synthesized in the nervous system can diffuse freely because of its small molecular structure and electrical neutrality, which makes it a good candidate to be involved into a form of diffusive intracellular communication \cite{Lancaster_1994}.

Voltage dependent $Ca^{2+}$-channels open due to a depolarization of the membrane potential in the course of an action potential, which causes a significant increase of the intracellular Calcium concentration (relative to the low concentration at rest) \cite[p.~98-100]{Hille_Ion_Channels}. A common theoretical description of voltage-dependent ion channels is provided by the Goldman-Hodgkin-Katz equations, see \cite[p.~445-451]{Hille_Ion_Channels}. As such, $Ca^{2+}$ constitutes a causal link between spiking activity and the production of NO. Experimental studies have also suggested that NO can act as an inhibitory diffusive signaling pathway, decreasing intrinsic excitability \cite{Steinert_NO_2011}, or generally suggest a role of NO in maintaining a functional state of activity \cite{Pape_NO}. In experiments upon neurons in the medial nucleus of the trapezoid body Steinert et al. found that cells that were subject to NO-signaling by an external source exhibited an increment in outward $\mathrm{K^+}$ current, which, due to its hyperpolarizing effect, can be regarded as a form of intrinsic inhibition.

@@

It has been generally argued that the set of rules used in these recurrent networks represent a realization of a minimum ensemble of functional compartments required in order to give rise to the desired capabilities. The network's ability of self-organization has not only been investigated in terms of network structure itself. It also successfully performed in unsupervised sequence-learning tasks, suggesting that the SORN can acquire and maintain associative memory \cite{Hartmann_2016}.

@@

400 excitatory LIF neurons and 80 inhibitory LIF neurons were assigned random positions across a square area of 1000 $\times$ 1000 $\mu m$. All but recurrent excitatory synapses were randomly generated before the start of the simulation until a desired connection fraction was reached. The connection probability between two neurons was calculated from a distant dependent Gaussian function with a standard deviation of 200 $\mu m$. For excitatory to inhibitory (EI) and inhibitory to excitatory (IE) synapses, the connection fraction was set to 0.1, and 0.5 for recurrent inhibitory synapses (II). These connections were kept at a fixed connection strength throughout the simulation. Furthermore, all synapses were simulated with a fixed (distance independent) conduction delay. See table \ref{syn_conn_params} for a summary of parameters.

Recurrent excitatory synapses were subject to a number of plastic mechanisms to be described in the following.

\begin{table}
\begin{tabular}{|l|l|l|l|l|}
\hline
\textbf{parameter} & \textbf{EE} & \textbf{EI} & \textbf{IE} & \textbf{II} \\ \hline
connection fraction & $\rightarrow 0.1$ & $0.1$ & $0.1$ & $0.5$ \\ \hline
initial connection strength & $0.0001 mV$ & $1.5 mV$ & $-1.5 mV$ & $-1.5 mV$ \\ \hline
conduction delay & $1.5 ms$ & $0.5 ms$ & $1.0 ms$ & $1.0 ms$ \\
\hline
\end{tabular}
\caption{Parameters of synaptic connections.}
\label{syn_conn_params}
\end{table}
\newpage
\subsubsection{Synaptic Plasticity}
\textbf{Synaptic Growth:} The random, distance dependent generation of new EE synapses was carried out n times once per second, where n is taken from a normal distribution with mean 920 and standard deviation $\sqrt{920}$. This constant growth rate was tuned to achieve the desired target concentration of 0.1 (see \ref{syn_conn_params}).

@@

\textbf{Spike Timing Dependent Plasticity:} An additive STDP rule was used as described e.g. in \cite{Zhang_STDP}. The change of weight between two neurons due to a pre- and postsynaptic spike (i $\rightarrow$ j) is defined as:
\begin{align}
\Delta w_{ji} &= \sum_k \sum_l W(t_j^l - t_i^k) \label{STDP_rule} \\
W(\Delta t) &= A_{+} exp(-\Delta t / \tau_{+}), & \Delta t > 0 \label{STDP_pos} \\
W(\Delta t) &= A_{-} exp(\Delta t / \tau_{-}), & \Delta t < 0 \label{STDP_neg}
\end{align}
Indexes k and l refer to the k-th and l-th pre- and postsynaptic spike respectively. Parameters were chosen to approximate data from \cite{Bi_Poo_STDP} and \cite{Froemke_STDP}, namely $\tau_{+} = 15 ms$, $A_{+} = 15 mV$, $\tau_{-} = 30 ms$ and $A_{-} = -7.5 mV$. However, we used the ``nearest neighbor" approximation for the sake of reduction of computational effort, only calculating the effect of the most recent pre-post pair of spikes for potentiation and post-pre pair for depression, yielding roughly the same value as the full summation due to the fast decay times $\tau_{+}$ and $\tau_{-}$ of the STDP-window.

@@

\textbf{Short Term Plasticity:} A short term plasticity (STP) mechanism acting on recurrent excitatory connections was implemented as presented in \cite{Markram_STP} as an additional stabilization of network activity. It modulates the effective synaptic weights by multiplying the value stored in the weight matrix $w_{ji}$ by two dynamic variables $x$ and $u$, $w^{effective}_{ji} = w_{ji}\cdot x \cdot u$, each synapse owning a pair $(x,u)$. The dynamics of these variables are given by:

>>>>

\begin{abstract}
Building upon on a self-organizing spiking neural network (LIF-SORN), we attempted to replace the intrinsic homeostatic control system used in earlier versions by a mechanism based on the diffusion of a neurotransmitter across the nervous tissue. The model of diffusive homeostasis was adopted from a paper by Sweeney et al. and models the tissue as a surface of square shape and a set of points on this surface, representing the positions of the neurons within the network. Excitatory neurons acted as a source of nitric oxide (NO), as well as as a sensor for the NO concentration at each individual position. Production and sensing of NO formed the basis of a feedback loop: Individual NO readouts caused an appropriate change within the intrinsic excitability of the neurons. The control system was closed by linking the rate of NO production to the neurons' firing rates.

The main goal of the modification of homeostasis was to allow for a broad and heavy tailed distribution of firing rates among the excitatory neural population. This feature of cortical activity has been extensively observed and studied experimentally, but could not be achieved by the formerly used single-cell homeostatic mechanism which bound firing rates of all neurons to a fixed target value. Indeed, our statistical analysis of spiking activity showed that diffusive homeostasis could significantly broaden the distribution of firing rates.

@@

\section{Introduction}
\subsection{Homeostasis in the Nervous System}
Among other functional roles of the mammalian central nervous system it is involved in maintaining crucial properties of the body in a functional regime. Some examples regarding the organism as a whole are e.g. body temperature and blood pressure, both being controlled by hypothalamus \cite[p. 484]{Bear_Exploring_the_Brain}. The term \textit{homeostasis} entailing these regulatory mechanisms was first introduced by Walter C. Cannon in 1932 \cite{Wisdom_of_the_Body_Cannon, Homeostasis_Queenan_2012}. In more recent years, experimental studies have found evidence for processes within the brain that do not function as a regulatory control for certain aspects of the body, but as a homeostatic control of neural activity itself \cite{Turrigiano_1994,Burrone_2003}. Many processes underlying this self-regulatory behavior have been found so far and many different biophysical interdependencies have shown to play a role. This diversity makes comparability of these mechanisms a rather difficult task. On a relatively general level though experiments have revealed two main types of neuronal regulation: synaptic homeostatic plasticity and intrinsic homeostasis \cite{Desai_2003}.

The notion of synaptic homeostatic plasticity subsumes the observation that neurons - being more or less active than a certain set point of mean activity - are regulated by means of their incoming synaptic weights. Theoretical and experimental studies have suggested that this type of regulation is taking place in a multiplicative manner, thereby retaining proportions between individual synaptic strengths \cite{Syn_Plast_Abbott}. One can regard this form of plasticity as an antagonist to other aspects of synaptic plasticity that affect synaptic strengths in an activity dependent, Hebbian fashion. Forms of regulation belonging to intrinsic homeostasis alter the neurons' capability of spiking upon external input by changing intrinsic properties of the nerve cell. More precisely, a neurons' ability to form an action potential can change by modifying the conductance of ion channels in its cell membrane \cite[p. 156]{Theor_Neur_Dayan}.

Theoretical models of homeostasis are generally based on a form of feedback control. This is also known as a \textit{closed loop control} in control theory. All these systems include a number of necessary functional components, which will be briefly introduced in the following. Concepts and terms were taken from \cite[p. 59-64]{Cybernetik_Systems_Cruse_2006}.
\newpage
\subsubsection{Theory of Control Systems} \label{Control_Sys_Theo_Section}
A closed loop control generally must contain some \emph{process} that is to be regulated. Despite its potential internal complexity, its functional role in the context of a feedback system is to simply generate a time-dependent output based on the history of a time-dependent input. This output is what one wishes to control and regulate by means of the feedback loop and thus represents the signal that is further processed. It might not be possible however to directly quantify this desired output in terms of some internal state of the process. The firing rate of a neuron for example is an abstract quantity, which can not be \emph{directly} related to some neuronal state variable. Consequently, it might be necessary to transform the actual output of the process into a different signal by a \textit{feedback transducer}. The (potentially transformed) output is fed into a \textit{comparator} in the next step, which---unsurprisingly---compares the given output of the process to some given target or \textit{reference}, typically by subtraction. Note that this reference can but does note have to be a time-dependent signal. The term \textit{homeostasis} however generally refers to a fixed reference, while implementations with a time dependent reference can be regarded as a form of \textit{servomechanism}. The difference between reference and the feedback transducer's signal is known as the \textit{error} signal. This error signal is given to a \textit{controller}, which transforms this error signal into an appropriate input signal for the initially introduced process. This is also the point in the control loop where \textit{external disturbances} come into play by being added to the input generated by the controller. We illustrated the described flow of feedback control in Figure \ref{Closed_Loop_Control_Illustration}.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/illustrations/closed_loop_control.eps}
\end{center}
\caption{Schematic illustration of a closed loop control as described in Section \ref{Control_Sys_Theo_Section}}
\label{Closed_Loop_Control_Illustration}
\end{figure}
Based on the way the controller generates input for the dynamic process based on the error signal one can classify control loops by three types:
\newpage
\begin{itemize}
\item A \textit{proportional controller} generates a signal that is proportional to the error signal it receives. The most important property of a proportional controller is that the stationary state of the control loop generally does not imply an error signal equal to zero.
\item An \textit{integrator controller} uses an integration over the past of the error signal to obtain its output. The stationary solution necessarily implies that the error signal is zero.
\item A \textit{differential controller} generates its output by means of the time derivative of the error signal. This type of controller can quickly react to momentary changes but does not compensate for sustained deviations of the error signal from zero.
\end{itemize}
It should be pointed out that these types of control mechanisms can be simultaneously used in a control system, e.g. through a linear combination. Furthermore, the presented terms refer to concepts of linear control theory. In the study of nonlinear feedback systems they thus only apply in a limited regime where a first order approximation of the system is sufficiently exact.

\subsubsection{Error Signals and Controllers of Neuronal Homeostasis}
A key step in characterizing a control loop in the context of biological systems should be to identify the previously described abstract elements. While the functional separation illustrated in Figure \ref{Closed_Loop_Control_Illustration} might seem intuitive and ``straight forward", it is not obvious which physiological element should be attributed to a certain abstract element.
As a start, we state that the process and its output controlled by homeostasis is the generation of action potentials and the resulting firing rate. Firing rate is---as stated earlier---an abstract quantity, requiring a feedback transducer to translate the neuron's spiking activity into a ``sensible" quantity. In a nerve cell this could be represented by e.g. the relation between the firing of action potentials and the temporary rise in $\mathrm{Ca^{2+}}$ concentration. An example for a homeostatic model making use of intracellular calcium concentration as a measure of spiking activity is the LeMasson-Abbott model \cite{LeMasson_1993}. More complex chains of activity encoding are of course possible and we will introduce a recent model making additional use of the concentration of nitric oxide in Section \ref{DIff_Hom_Intro_Section}. Given a representation of activity, its deviation from a reference value will cause other physiological quantities to affect the neuron's ability to generate action potentials. Referring to the concepts introduced in the previous section one could then ask what type of control best resembles the actual neuronal control. The fact that neurons, whose firing rate was artificially increased or suppressed over a long time, can reliably tune their activity back to the initial state (see e.g. \cite{Burrone_2002}) rules out a purely differential type of control, and rather points to an integral or proportional control. It is clear however that the ability of a biological system to encode the accumulation of an error signal is limited by the finite amount of available resources. The conductance of an ion channel for example has natural upper and lower bounds that it can not surpass. We will pick up this issue in Section \ref{Possible_Modifications_Section}.

\subsubsection{The Role of Firing Rate Heterogeneity---a Dichotomy between Heterogeneity and Control?}
Modeling homeostasis necessarily includes a form of error estimation, which requires setting some predefined target that the feedback control will attempt to achieve. Of course this raises the question of how to choose these homeostatic targets. Experimental studies have found that neuronal populations exhibit a strong heterogeneity within their individual activity. Many statistical analyses found particular evidence for a log-normal like distribution of firing rates \cite{Buzsaki_Fir_Rates_2014,Wohrer_Fir_Rates_2012}. Following these results, one approach of modeling this feature would be to randomly distribute individual targets according to the aforementioned statistics. A problem of this procedure is the absence of any particular deeper physiological motivation or causation. In the light of this conceptual problem, one could, due to a lack of better knowledge, simply define a single global target of activity.

However, numerous experimental and theoretical studies have suggested that the presence of both slow- and fast-firing cells is not to be regarded as an ignorable, mere side effect of brain dynamics \cite{Buzsaki_2004,Marsat_2010,Tripathy_2013}. It rather seems to be that skewed and heavy-tailed firing rate statistics are necessary with respect to the preservation of stable synaptic structures, being related to long-term memory, while still allowing synaptic rewiring in order to adapt to changes in external stimuli. For example, experiments by Dragoi et al. have found that during remapping of hippocampal place cells, neurons with low firing rates are subject to stronger changes within their field of spatial representation than those with high firing rates \cite{Dragoi_2003}. In this context Buzsáki and Mizuseki argue that a small subset of highly active neurons may serve to form relatively strongly connected and stable sub-networks that can quickly react to momentary changes in the environment, while fine tuned adaptation involves the slower firing majority of cells \cite{Buzsaki_Fir_Rates_2014}. A realistic network model of neural activity and plasticity should therefore allow for a broad distribution of firing rates to cope with these functional aspects.

@@

Another problem that one might consider is the fact that temporal fluctuations of activity are believed to be a major part of information transmission and processing within the brain \cite[p. 123--150]{Theor_Neur_Dayan}. Homeostasis as described above could potentially override information by eliminating these fluctuations of activity. However, it is argued that homeostatic mechanisms act on time scales of hours to days and thus do not interfere with changes in activity taking place on the order of seconds or below \cite{Turrigiano_2011}. Returning to the previous question, though the slowness of homeostatic adaptation might allow for temporal fluctuations of activity, it does not overcome the conceptual dichotomy between single cell homeostasis and persistent variability in spontaneous activity. Marder and Goaillard speculated that instead of a mechanism controlling activity on a single-cell level, one might consider the possibility of a ``global sensor of network performance to provide error signals for tuning of the network" \cite{Marder_2006}. This concept was used by Sweeney et al. as the basic idea for the introduction of a diffusive, intercellular homeostatic mechanism, which will be introduced in the next section.

\subsection{Diffusive Homeostasis}\label{DIff_Hom_Intro_Section}
In this thesis we examined \textit{diffusive homeostasis} as a possible candidate to replace a previously used model of single-cell intrinsic plasticity. The idea and modeling of diffusive homeostasis was adopted from a paper by Sweeney et al. \cite{Sweeney_Paper}, which models neural tissue as a two-dimensional surface and a set of points representing the neurons' positions. The group of excitatory neurons acted as a point-source of nitric oxide (NO) as well as as a sensor for the NO-concentration at each individual position. The production and sensing of NO forms the basis of a feedback loop: The individual NO-readout is fed into a comparator which causes an appropriate change within the internal firing threshold of the neuron, in turn altering the neuron's firing rate. The control system is then closed by linking the rate of NO-production to the neuron's firing rate. The results by Sweeney et al. suggest that diffusive signaling could resolve the dichotomy between overall stability of firing activity and the need to allow flexibility among individual neurons' activities. Through the diffusive signal, each neurons receives---intuitively speaking---a mixture of its own activity and its neighboring neurons. Individual tuning of firing rates is thereby suppressed while the overall population activity is kept at a constant level.

\subsubsection{Nitric Oxide in the Nervous System}\label{NO_Experiments_Section}
The model presented in \cite{Sweeney_Paper} includes spatial interaction across excitatory neurons through the diffusion of nitric oxide (NO). Enzymes that are responsible for NO synthesis (NOS) are present in different areas of the body, namely being involved in the dilation of blood vessels (endothelial NOS, eNOS), the immune system (inducible NOS, iNOS) and the CNS (neuronal NOS, nNOS) \cite{NOS_Mammals}. These Enzymes differ in their functionality and dependence on the presence of other molecules. In particular, nNOS is sensitive to the concentration of $\mathrm{Ca^{2+}}$ ionic Calcium \cite{Knowles_Ca_nNOS,Steinert_NO}. Nitric oxide that was synthesized in the nervous system can diffuse freely because of its small molecular structure and electrical neutrality, which makes it a good candidate to be involved into a form of diffusive intercellular communication \cite{Lancaster_1994}.

Voltage dependent $Ca^{2+}$-channels open due to a depolarization of the membrane potential in the course of an action potential, which causes a significant increase of the intracellular Calcium concentration (relative to the low concentration at rest) \cite[p.~98--100]{Hille_Ion_Channels}. A common theoretical description of voltage-dependent ion channels is provided by the Goldman-Hodgkin-Katz equations, see \cite[p.~445--451]{Hille_Ion_Channels}. As such, $Ca^{2+}$ constitutes a causal link between spiking activity and the production of NO. Experimental studies have also suggested that NO can act as an inhibitory diffusive signaling pathway, decreasing intrinsic excitability \cite{Steinert_NO_2011}, or generally suggest a role of NO in maintaining a functional state of activity \cite{Pape_NO}. In experiments on neurons in the medial nucleus of the trapezoid body Steinert et al. found that cells that were subject to NO-signaling by an external source exhibited an increment in outward $\mathrm{K^+}$ current, which, due to its hyperpolarizing effect, can be regarded as a form of intrinsic inhibition \cite{Steinert_NO}.

@@

It has been generally argued that the set of rules used in these recurrent networks represents a realization of a minimum ensemble of functional elements required in order to give rise to the desired capabilities. The network's ability of self-organization has not only been investigated in terms of network structure itself. It also successfully performed in unsupervised sequence-learning tasks, suggesting that the SORN can acquire and maintain associative memory \cite{Hartmann_2016}.

@@

400 excitatory LIF neurons and 80 inhibitory LIF neurons were assigned random positions across a square area of 1000 $\times$ 1000 $\mu m$. All but recurrent excitatory synapses were randomly generated before the start of the simulation until a desired connection fraction was reached. The connection probability between two neurons was calculated from a distance-dependent Gaussian function with a standard deviation of 200 $\mu m$. For excitatory to inhibitory (EI) and inhibitory to excitatory (IE) synapses, the connection fraction was set to 0.1, and 0.5 for recurrent inhibitory synapses (II). These connections were kept at a fixed connection strength throughout the simulation. Furthermore, all synapses were simulated with a fixed (distance independent) conduction delay. See table \ref{syn_conn_params} for a summary of parameters.

Recurrent excitatory synapses were subject to a number of plastic mechanisms to be described in Section \ref{Section_Syn_Plast}.

\begin{table}
\begin{tabular}{|l|l|l|l|l|}
\hline
\textbf{parameter} & \textbf{EE} & \textbf{EI} & \textbf{IE} & \textbf{II} \\ \hline
connection fraction & $\rightarrow 0.1$ & $0.1$ & $0.1$ & $0.5$ \\ \hline
initial connection strength & $0.0001 mV$ & $1.5 mV$ & $-1.5 mV$ & $-1.5 mV$ \\ \hline
conduction delay & $1.5 ms$ & $0.5 ms$ & $1.0 ms$ & $1.0 ms$ \\
\hline
\end{tabular}
\caption{Parameters of synaptic connections.}
\label{syn_conn_params}
\end{table}
\newpage

\subsubsection{Neuron Model}
We used a leaky integrate-and-fire model for all neurons in the network, whose dynamics are described by a stochastic differential equation:
\begin{equation}
{\tau_m}dV = -(V-E_l)dt + \sqrt{\tau_m} \sigma dW
\label{LIF_Dynamics}
\end{equation}
where $V$ is the membrane potential, $E_l$ is the equilibrium membrane potential, $\tau_m$ is the time constant of the membrane, $\sigma$ is the standard deviation of the noise term and $dW$ is the standard Wiener process. A neuron is said to spike when its membrane potential reaches the threshold voltage $V_t$. The voltage is then reset to $V_r$. A refractory period was not implemented. A presynaptic spike causes a simple (delayed, see Table \ref{syn_conn_params}) increment of the membrane potential of the postsynaptic neuron by $w^{\rm effective}_{ji}$. Table \ref{LIF_neuron_params} summarizes the aforementioned set of parameters.
\begin{table}[H]
\begin{tabular}{|l|l|l|}
\hline
\textbf{parameter} & \textbf{exc. neur.} & \textbf{inh. neur.}\\ \hline
$\mathrm{E_l}$ & $\mathrm{-60\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\tau_m}$ & $\mathrm{20\;ms}$ & $\mathrm{20\;ms}$ \\ \hline
$\mathrm{V_r}$ & $\mathrm{-70\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\sigma}$ & $\mathrm{\sqrt{5}\;mV}$ & $\mathrm{\sqrt{5}\;mV}$ \\ \hline
$\mathrm{V_t}$ & subject to IP & $\mathrm{-58\;mV}$ \\ 
\hline
\end{tabular}
\caption{Parameters of LIF neuron}
\label{LIF_neuron_params}
\end{table}

\subsubsection{Synaptic Plasticity}\label{Section_Syn_Plast}
\textbf{Synaptic Growth:} Once per second, $\mathrm{n}$ new EE synapses were generated, where n is taken from a normal distribution with mean 920 and standard deviation $\sqrt{920}$. This constant growth rate was tuned to achieve the desired target concentration of 0.1 (see \ref{syn_conn_params}).

@@

\textbf{Spike Timing Dependent Plasticity:} An additive STDP rule was used as described e.g. in \cite{Zhang_STDP}. The change of weight between two neurons due to a pre- and postsynaptic spike (i $\rightarrow$ j) is defined as:
\begin{align}
\Delta w_{ji} &= \sum_k \sum_l W(t_j^l - t_i^k) \label{STDP_rule} \\
W(\Delta t) &= A_{+} \mathrm{exp}(-\Delta t / \tau_{+}), & \Delta t \geq 0 \label{STDP_pos} \\
W(\Delta t) &= A_{-} \mathrm{exp}(\Delta t / \tau_{-}), & \Delta t < 0 \label{STDP_neg}
\end{align}
Indexes k and l refer to the k-th and l-th pre- and postsynaptic spike respectively. Parameters were chosen to approximate data from \cite{Bi_Poo_STDP} and \cite{Froemke_STDP}, namely $\tau_{+} = 15 ms$, $A_{+} = 15 mV$, $\tau_{-} = 30 ms$ and $A_{-} = -7.5 mV$. However, we used the ``nearest neighbor" approximation for the sake of reduction of computational effort, only calculating the effect of the most recent pre-post pair of spikes for potentiation and post-pre pair for depression, yielding roughly the same value as the full summation due to the fast decay times $\tau_{+}$ and $\tau_{-}$ of the STDP-window.

@@

\textbf{Short Term Plasticity:} A short term plasticity (STP) mechanism acting on recurrent excitatory connections was implemented as presented in \cite{Markram_STP} as an additional stabilization of network activity. It modulates the effective synaptic weights by multiplying the value stored in the weight matrix $w_{ji}$ by two dynamic variables $x$ and $u$, $w^{\rm effective}_{ji} = w_{ji}\cdot x \cdot u$, each synapse owning a pair $(x,u)$. The dynamics of these variables are given by:
